---
title: "Untitled"
output: html_document
date: "2022-11-02"
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r setup, include = FALSE}
pacman::p_load(tidyverse, tximport, biomaRt, DESeq2, WGCNA, cowplot, patchwork)
```

```{r cliff}
# prep metadata -----------------------------------------------------------

metadata = data.frame(names = list.dirs("./data/kallisto",full.names = F),
                      files = list.dirs("./data/kallisto",full.names = T))[-1,]
metadata$files = paste0(metadata$files,"/abundance.h5")
metadata$time = gsub(".*_(T.*?)_.*","\\1",metadata$names)
metadata$background = gsub(".*?_(.*?)_.*","\\1",metadata$names)
metadata$treatment = gsub(".*_T.*?_(.*?)_.*","\\1",metadata$names)
metadata$treatment[metadata$time=="T0"] = "uninf"
metadata$group = paste0(metadata$background, "___",metadata$treatment, "___", metadata$time)
rownames(metadata) = metadata$names

# import data -------------------------------------------------------------
files = metadata$files
names(files) = metadata$names
tx = tximport::tximport(files, type = "kallisto", txOut = TRUE)

# Build reference mapping -------------------------------------------------
mart = useEnsembl(biomart = "ensembl",
               dataset = "mmusculus_gene_ensembl",
               mirror = "useast")

t2g = getBM(attributes = c("refseq_mrna", "external_gene_name"),mart = mart)
t2g = dplyr::rename(t2g, 
                    target_id = refseq_mrna,
                    gene_id = external_gene_name)
t2g = t2g[t2g$target_id!="",]
t2g = t2g[match(gsub("\\..*","",rownames(tx$abundance)),t2g$target_id),]
t2g$target_id = rownames(tx$abundance)
t2g = t2g[!is.na(t2g$gene_id),]

gene = summarizeToGene(tx,tx2gene = t2g)
# provides identical result to above 
# gene <- tximport(files, type = "kallisto", tx2gene = t2g)

gene_lengthScaledTPM = summarizeToGene(tx,tx2gene = t2g, countsFromAbundance = "lengthScaledTPM")
```

```{r metadata}
metadata <- tibble::tribble(
                 ~name, ~background, ~time, ~treatment, ~replicate,
            "01_wt_T0_1",      "wt",  "T0",    "uninf",         "1",
            "02_wt_T0_2",      "wt",  "T0",    "uninf",         "2",
            "03_wt_T0_3",      "wt",  "T0",    "uninf",         "3",
         "04_NLRP3_T0_1",   "NLRP3",  "T0",    "uninf",         "1",
         "05_NLRP3_T0_2",   "NLRP3",  "T0",    "uninf",         "2",
         "06_NLRP3_T0_3",   "NLRP3",  "T0",    "uninf",         "3",
          "07_MAVS_T0_1",    "MAVS",  "T0",    "uninf",         "1",
          "08_MAVS_T0_2",    "MAVS",  "T0",    "uninf",         "2",
          "09_MAVS_T0_3",    "MAVS",  "T0",    "uninf",         "3",
       "10_wt_T24_inf_1",      "wt", "T24",      "inf",         "1",
       "11_wt_T24_inf_2",      "wt", "T24",      "inf",         "2",
       "12_wt_T24_inf_3",      "wt", "T24",      "inf",         "3",
     "13_wt_T24_uninf_1",      "wt", "T24",    "uninf",         "1",
     "14_wt_T24_uninf_2",      "wt", "T24",    "uninf",         "2",
     "15_wt_T24_uninf_3",      "wt", "T24",    "uninf",         "3",
    "16_NLRP3_T24_inf_1",   "NLRP3", "T24",      "inf",         "1",
    "17_NLRP3_T24_inf_2",   "NLRP3", "T24",      "inf",         "2",
    "18_NLRP3_T24_inf_3",   "NLRP3", "T24",      "inf",         "3",
  "19_NLRP3_T24_uninf_1",   "NLRP3", "T24",    "uninf",         "1",
  "20_NLRP3_T24_uninf_2",   "NLRP3", "T24",    "uninf",         "2",
  "21_NLRP3_T24_uninf_3",   "NLRP3", "T24",    "uninf",         "3",
     "22_MAVS_T24_inf_1",    "MAVS", "T24",      "inf",         "1",
     "23_MAVS_T24_inf_2",    "MAVS", "T24",      "inf",         "2",
     "24_MAVS_T24_inf_3",    "MAVS", "T24",      "inf",         "3",
   "25_MAVS_T24_uninf_1",    "MAVS", "T24",    "uninf",         "1",
   "26_MAVS_T24_uninf_2",    "MAVS", "T24",    "uninf",         "2",
   "27_MAVS_T24_uninf_3",    "MAVS", "T24",    "uninf",         "3"
  ) 

# creating treatment2 from a linear combination of time and treatment
metadata <- metadata |>
mutate(time_treat = case_when(time == "T24" & treatment == "uninf" ~ "bystander",
                              time == "T24" & treatment ==   "inf" ~  "infected",
                              time ==  "T0" & treatment == "uninf" ~ "baseline"),
       back_treat =  paste0(background, "_", treatment)
       )

metadata <- metadata |>
  column_to_rownames("name") |>
  mutate(background = background |> as.factor() |> relevel(ref = "wt"),
         time = time |> as.factor() |> relevel(ref = "T0"),
         treatment = treatment |> as.factor() |> relevel(ref = "uninf"),
         time_treat = time_treat |> as.factor(),
         time_treat = fct_relevel(time_treat, c("baseline", "bystander", "infected")),
         back_treat = back_treat |> as.factor() |> relevel(ref = "wt_uninf")
         ) |> 
  dplyr::select(background, time, treatment, time_treat, back_treat, replicate)
```

## what is the effect of including an extra term in the model

-   an extra additive term in the model changes the underlying test from an unpaired t-test to a paired t-test.
-   it does not change the slope estimate (beta), it only changes its SE, and hence the t- & p-value
-   Quote from [@Sainani2010] explaining why the SE changes, "... the paired t-test only has to account for one source of variability (variability within pairs) rather than 2 sources (variability from two groups of twins)."
-   The unpaired t-test uses the third formula but the paired t-test uses the fourth formula of the [five formulas given here](https://miroslavtushev.medium.com/a-simple-trick-to-understand-the-t-test-2c2a9e7f1dc5)
-   Restated in terms of the metadata here, the SE changes because an unpooled rather than a pooled variance is calculated. That unpooled variance is calculated over the $(un)infected-baseline$ pairwise differences (where the differences are paired within the respective backgrounds).

## what does the intercept include

-   the intercept includes first levels of both the first and the second terms.
-   removing the intercept still presents the estimates for all levels of the second term relative to its first level.

## what is the meaning of interactions

-   An interaction represents the linear combination of two preexisting terms. In a model without interaction, all the lines have the same fixed value; but in a model with interaction they are free to vary.

## What information does dropping a term from a multiple linear regression model provide?

- By dropping a term from the model and comparing the fit statistics (e.g., R-squared, adjusted R-squared, residual standard error) of the reduced model to the original model, we can determine whether the dropped term was contributing significantly to the predictive power of the model.

## Consider a multiple regression model with two two-level categorical predictors and an interaction term (the third term in it is the interaction term). What is the relationship between the independent effect, the incremental effect, and the estimated marginal means of the second term?

The relationship between the independent effect, the incremental effect, and the estimated marginal means of the second term depends on the values of the other terms in the model. The independent effect of the second term is the change in the expected value of the dependent variable when the second term changes by one unit, holding the other terms constant. The incremental effect of the second term is the change in the expected value of the dependent variable when the second term changes by one unit, regardless of the values of the other terms. The estimated marginal means of the second term are the predicted values of the dependent variable for each level of the second term, averaged over all other terms in the model [1](https://stats.stackexchange.com/questions/577537/interpreting-main-effect-with-significant-interaction-term-in-continuous-by-cont).

In a model with two two-level categorical predictors and an interaction term, such as:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2 + \epsilon$$

where $x_1$ and $x_2$ are dummy variables that take values 0 or 1, the independent effect of $x_2$ is $\beta_2$ when $x_1 = 0$, and $\beta_2 + \beta_3$ when $x_1 = 1$. The incremental effect of $x_2$ is $\beta_2 + \frac{\beta_3}{2}$, which is a weighted average of its independent effects at different levels of $x_1$. The estimated marginal means of $x_2$ are $\beta_0 + \frac{\beta_1}{2}$ when $x_2 = 0$, and $\beta_0 + \frac{\beta_1}{2} + \beta_2 + \frac{\beta_3}{2}$ when $x_2 = 1$. These are also weighted averages of its predicted values at different levels of $x_1$ [2](https://cran.r-project.org/web/packages/sjPlot/vignettes/plot_marginal_effects.html).

Therefore, to interpret the effect of $x_2$, you need to consider both its main effect and its interaction effect with $x_1$. You can also plot the marginal effects or use confidence intervals to assess the significance and magnitude of the effect [3](http://www.sthda.com/english/articles/40-regression-analysis/164-interaction-effect-in-multiple-regression-essentials/).

The estimated marginal means of $x_1$ are $\beta_0 + \frac{\beta_2}{2}$ when $x_1 = 0$, and $\beta_0 + \frac{\beta_2}{2} + \beta_1 + \frac{\beta_3}{2}$ when $x_1 = 1$. These are also weighted averages of its predicted values at different levels of $x_2$.

## when using deseq2 with multiple terms, how can you calculate the se for a contrast?

The standard error for a contrast will be calculated by DESeq2 using the Wald test. The Wald test compares the estimated coefficient for the contrast to zero, divided by its standard error. The standard error for the contrast is computed from the variance-covariance matrix of the coefficients, which takes into account the correlation between them. 

```{r testing, eval=FALSE, include=FALSE}
metadata2  <- metadata |> filter(background != "NLRP3", time_treat != "infected")

metadata2$abundance <-  ba_by |>  counts(normalized=TRUE) %>% `[`(3, ) |> log2()

with_inter_m <- lm(abundance ~ background * time_treat, 
                      data = metadata2 |> filter(background != "NLRP3")
                      )

without_inter_m <- lm(abundance ~ background + time_treat, 
                      data = metadata2 |> filter(background != "NLRP3")
                      )

emmeans::emmeans(with_inter_m, background ~ time_treat | background, infer = TRUE)

with_inter <- emmeans::emmip(with_inter_m, background ~ time_treat, CIs = TRUE) + 
  labs(title = "With Interaction", y = "Gene '1' abundance") +
  ggrepel::geom_text_repel(data = function(x) subset(x,time_treat == "infected"),
                           aes(label = background)) +
  theme_half_open() + 
  theme(axis.title.x = element_blank(),
        legend.position = "none")

without_inter <- emmeans::emmip(without_inter_m, background ~ time_treat, CIs = F) + 
  labs(title = "Without Interaction", y = "Gene '1' abundance") +
  ggrepel::geom_text_repel(data = function(x) subset(x,time_treat == "infected"),
                           aes(label = background)) +
  theme_half_open() + 
  theme(axis.title.x = element_blank(),
        legend.position = "none")

with_inter + without_inter
```

1.  There is no correlation between the samples with the same replicate number across different genotypes. In other words: the replicates are nested within genotypes, not across them
2.  How to specify nesting <https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#model-specification>
3.  Does comparing models with differing terms provide their independent effect? If so, get only the independent effect of treatment or independent effect of genotype:treatment interaction, ignore the independent effect of genotype.
4. Rank genes on the basis of their enrichment scores.

```{r deg}
ba_by <- DESeqDataSetFromMatrix(
  countData = gene$abundance |> as.data.frame() |> dplyr::select(-contains(c("NLRP3", "T24_inf"))) |> round(),
  colData = metadata |> filter(background != "NLRP3", time_treat != "infected"),
  design = ~ background * time_treat
  ) |> DESeq()

ba_inf <- DESeqDataSetFromMatrix(
  countData = gene$abundance |> as.data.frame() |> dplyr::select(-contains(c("NLRP3", "T24_uninf"))) |> round(),
  colData = metadata |> filter(background != "NLRP3", time_treat != "bystander"),
  design = ~ background * time_treat
  ) |> DESeq()

by_inf <- DESeqDataSetFromMatrix(
  countData = gene$abundance |> as.data.frame() |> dplyr::select(-contains(c("NLRP3", "T0"))) |> round(),
  colData = metadata |> filter(background != "NLRP3", time_treat != "baseline"),
  design = ~ replicate + background * time_treat
  ) |> DESeq()

# # Generate results object
# resultsNames(ba_by)

# by default results outputs the last named contrast above

#####
ba_by_intercept <- results(name = "Intercept", object = ba_by) |> 
  as.data.frame() |> rownames_to_column()

ba_by_background <- results(name = "background_MAVS_vs_wt", object = ba_by) |> 
  as.data.frame() |> rownames_to_column()

ba_by_treat <- results(name = "time_treat_bystander_vs_baseline", object = ba_by) |> 
  as.data.frame() |> rownames_to_column()

ba_by_interaction <- results(name = "backgroundMAVS.time_treatbystander", object = ba_by) |>
  as.data.frame() |> rownames_to_column()

#####
ba_inf_intercept <- results(name = "Intercept", object = ba_inf) |> 
  as.data.frame() |> rownames_to_column()

ba_inf_background <- results(name = "background_MAVS_vs_wt", object = ba_inf) |> 
  as.data.frame() |> rownames_to_column()

ba_inf_treat <- results(name = "time_treat_infected_vs_baseline", object = ba_inf) |> 
  as.data.frame() |> rownames_to_column()

ba_inf_interaction <- results(name = "backgroundMAVS.time_treatinfected", object = ba_inf) |>
  as.data.frame() |> rownames_to_column()

#####
by_inf_intercept <- results(name = "Intercept", object = by_inf) |> 
  as.data.frame() |> rownames_to_column()

by_inf_background <- results(name = "background_MAVS_vs_wt", object = by_inf) |> 
  as.data.frame() |> rownames_to_column()

by_inf_treat <- results(name = "time_treat_infected_vs_bystander", object = by_inf) |> 
  as.data.frame() |> rownames_to_column()

by_inf_interaction <- results(name = "backgroundMAVS.time_treatinfected", object = by_inf) |>
  as.data.frame() |> rownames_to_column()
```

```{r emm}
head(coef(ba_by)) |> as.data.frame() |> rownames_to_column() |> dplyr::slice(3) |> 
  # rename("Intercept" = "WT_baseline", 
  #        "background_MAVS_vs_wt" = "MAVS_baseline", 
  #        "time_treat_bystander_vs_baseline" = "WT_bystander", 
  #        "backgroundMAVS.time_treatbystander" = "MAVS_bystander") |> 
  transmute(
    Gene = rowname,
    
    WT_baseline = Intercept + (time_treat_bystander_vs_baseline)/2,
    
    MAVS_baseline = Intercept + (background_MAVS_vs_wt)/2,
    
    WT_bystander = Intercept  + background_MAVS_vs_wt + (time_treat_bystander_vs_baseline)/2 +
      (backgroundMAVS.time_treatbystander)/2,
    
    MAVS_bystander = Intercept + (background_MAVS_vs_wt)/2 + time_treat_bystander_vs_baseline +
      (backgroundMAVS.time_treatbystander)/2
    
    )|>
  pivot_longer(!Gene, names_to = "group", values_to = "emm") |> 
  mutate(
    background = str_split_fixed(string = group, pattern = "_", n = 2) |> as.data.frame() |> pluck("V1"),
    time_treat = str_split_fixed(string = group, pattern = "_", n = 2) |> as.data.frame() |> pluck("V2")
  ) |> 
  ggplot(aes(x = time_treat, y = emm, group = background, color = background)) + 
  geom_line() + 
  geom_point() +
  ggrepel::geom_text_repel(data = function(x) subset(x, time_treat == "bystander"), aes(label = background)) +
  theme_half_open() +
  theme(axis.title.x = element_blank(), legend.position = "none") +
  labs(y = "Gene '1' abundance")

head(coef(ba_by, SE = TRUE)) |> as.data.frame() |> rownames_to_column() |> dplyr::slice(3) |> 
  # rename("Intercept" = "WT_baseline", 
  #        "background_MAVS_vs_wt" = "MAVS_baseline", 
  #        "time_treat_bystander_vs_baseline" = "WT_bystander", 
  #        "backgroundMAVS.time_treatbystander" = "MAVS_bystander") |> 
  transmute(
    Gene = rowname,
    WT_baseline = SE_Intercept,
    MAVS_baseline = SE_Intercept  + SE_background_MAVS_vs_wt,
    WT_bystander = SE_Intercept  + SE_time_treat_bystander_vs_baseline,
    MAVS_bystander = SE_WT_bystander  + SE_backgroundMAVS.time_treatbystander)

```

# GSEA

Quote from fgsea reference manual describing the columns of the output. <https://bioconductor.org/packages/release/bioc/manuals/fgsea/man/fgsea.pdf>

Each row corresponds to a tested pathway. The columns are the following: • pathway -- name of the pathway as in 'names(pathway)'; • pval -- an enrichment p-value; • padj -- a BH-adjusted p-value; • ES -- enrichment score, same as in Broad GSEA implementation; • NES -- enrichment score normalized to mean enrichment of random samples of the same size; • nMoreExtreme' -- a number of times a random gene set had a more extreme enrichment score value; • size -- size of the pathway after removing genes not present in 'names(stats)'. • leadingEdge -- vector with indexes of leading edge genes that drive the enrichment, see <http://software.broadinstitute.org/gsea/doc/GSEAUserGuideTEXT.htm#_Running_a_Leading>

```{r gene_sets}
library(fgsea)
library(dorothea)
library(progeny)

hallmarks <- fgsea::gmtPathways("./data/pathway datasets/hallmark.genesets.v6.1.symbols.gmt") #50 gene sets
kegg <- fgsea::gmtPathways("./data/pathway datasets/kegg.genesets.v6.1.symbols.gmt") #186
go <- fgsea::gmtPathways("./data/pathway datasets/GOTerms.BP.v6.1.symbols.gmt") #4436
reactome <- fgsea::gmtPathways("./data/pathway datasets/reactome.genesets.v6.1.symbols.gmt") #674

gene_sets <- c(hallmarks, kegg, go, reactome)
```

```{r cliff_gsea}
mtb <- read_csv("~/mtb/results/DESeq/logFCs.csv")

gene_names <- mtb |> drop_na() |> pluck("...1")

for(i in 2:73){ 
  names(mtb[[i]]) <- gene_names

}

fold_changes <- vector("list", length = 72)

for(i in 2:73){
  fold_changes[[i]] <- mtb[[i]] |> sort()
}

pathways <- vector("list", length = 72)

for(i in 2:28){
  pathways[[i]] <- fgsea(pathways = gene_sets,
                         stats = fold_changes[[i]],
                         minSize=5,
                         maxSize=5000,
                         nproc = 2)
}

names(pathways) <-  mtb |> dplyr::select(-1) |> names() |> str_replace_all("___", "_") |> str_replace_all(" -- vs -- ", "—")

singletons <- names(pathways) |> str_split_fixed("—", n =2) |> as.data.frame() |> pluck(1) |> unique()

keep <- map_dfr(.x = 1:8, ~expand.grid(singletons[.x], singletons[(.x+1):9])) |> transmute(Var3 = paste0(Var1, "—", Var2)) |> pluck(1)

pathways2 <- pathways[keep]

map_dfr(.x = pathways2 |> discard(is.null), ~dplyr::filter(.x, pval <= 0.05), .id = "comparison") |> 
write.xlsx(file = "pathways1.xlsx")

write.xlsx(pathways2, file = "pathways2.xlsx")
```

Use the Bioconductor package ReportingTools for displaying the results [@Huntley2013]. Excerpt from @Love2015 about ReportingTools:

> ReportingTools will automatically generate dynamic HTML documents, including links to external databases using gene identifiers and boxplots summarizing the normalized counts across groups.

```{r}
# BiocManager::install("ReportingTools")
library("ReportingTools")
htmlRep <- HTMLReport(shortName="report", title="My report",
                      reportDirectory="./report")
publish(df |> slice_min(pval, n = 100), htmlRep)
url <- finish(htmlRep)
browseURL(url)
```
